---
title: "CoMEt"
author: "Héctor Climente"
date: "10/07/2015"
output:
  md_document:
    variant: markdown_github
---

We wanted to use [CoMEt](https://github.com/raphael-group/comet) to find cases of mutual exclusion without a priori knowledge of genesets.

```{r environment, include=FALSE}

source("~/SmartAS/Pipeline/scripts/variablesAndFunctions.r")
library(knitr)

```

# About CoMEt

[CoMEt](http://www.genomebiology.com/2015/16/1/160). There are some parameters that have to be decided by the user.

* Geneset size:
* Output:
  * Weight: coverage of a set M of k genes minus the coverage overlap (co-occurring mutations) of M.
  * Phi (score): result of an exact test of mutual exclusivity.

# Data preparation

```{r prepraration}

candidatesDf <- read.delim(paste0(workingDir,"/switches/tables/candidateList_splitByTumor_models_notNoise.txt"))

# pick functional ones
candidatesDf <- candidatesDf[candidatesDf$IsRelevant == 1,]

for (cancer in cancerTypes){
  cancer.df <- candidatesDf[candidatesDf$Tumor==cancer,c("GeneId","Patients_affected","Symbol")]
  
  genes_and_patients <- by(cancer.df,cancer.df$GeneId,function(x){
    p <- intersect(unlist(strsplit(as.character(x$Patients_affected[1]),",")),commonPatients)
    if (length(p)>0) {
      data.frame("Patient"=p,"Gene"=x$Symbol[1])
    }
  })
  genes_and_patients <- do.call("rbind",genes_and_patients)
  
  # create input for CoMEt
  alt_matrix <- ddply(genes_and_patients,.(Patient),summarise,"Switches"=toString(Gene))
  alt_matrix$Switches <- gsub(", ","\t",alt_matrix$Switches)
  
  write.table(alt_matrix,file=paste0(workingDir,"/switches/tables/",cancer,"_comet_onlySwitches.txt"),sep="\t", row.names=F, col.names=F, quote=F)
}
```

The result is a matrix where each patient is in a row and the switched genes are tab-separated.

# Run CoMEt

```{r run }

cancer <- "brca"
genesetSize <- 4

cometCommand <- paste("python","~/SmartAS/Pipeline_experimental/libs/comet/run_comet.py","-m", paste0(workingDir,"/switches/tables/",cancer,"_comet_onlySwitches.txt","-o",paste0("comet_",cancer),"-ks",genesetSize,sep=" ")

system(cometCommand)

output <- paste0("comet_",cancer,".sum.k",genesetSize,".1K.1.tsv")


```

```{r mutation_handle, eval=FALSE, include=FALSE}

candidatesDf <- read.delim(paste0(workingDir,"/switches/tables/candidateList_splitByTumor_models_notNoise.txt"))

# pick functional ones
candidatesDf <- candidatesDf[candidatesDf$IsRelevant == 1,]

allPatients_byCancerType <- ddply(candidatesDf,.(Tumor),summarise,patients=list(unique(unlist(strsplit(as.character(Patients_affected),",")))))

cancer <- "brca"

# read mutations
mutations <- read.delim(paste0(workingDir,"/Data/TCGA/Rawdata/",cancer,"_gene_mutation-functional-count_full.txt"),header=F)

# remove not assigned to patient and pcik genes and patients columns
mutations <- mutations[mutations$V10!=".",c("V4","V10")]
colnames(mutations) <- c("Gene","Patient")

patient_info <- do.call("rbind",strsplit(as.character(mutations$Patient),";"))
colnames(patient_info) <- c("Patient","Alteration")

gene_info <- do.call("rbind",strsplit(as.character(mutations$Gene),"|",fixed=T))
colnames(gene_info) <- c("Symbol","GeneId")

mutations_f <- as.data.frame(cbind(gene_info,patient_info))

mutations_f$Patient <- substr(mutations_f$Patient,1,nchar(as.character(mutations_f$Patient))-1)

# intersect the patients with any switch and the patients with any mutation
commonPatients <- intersect(allPatients_byCancerType$patients[allPatients_byCancerType$Tumor==cancer][[1]],mutations_f$Patient)

patients <- by(candidatesDf[candidatesDf$Tumor==cancer,c("GeneId","Patients_affected")],candidatesDf$GeneId[candidatesDf$Tumor==cancer],function(x){
  p <- intersect(unlist(strsplit(as.character(x[1,2]),",")),commonPatients)
  if (length(p)>0) {
    data.frame("Patient"=p,"GeneId"=x[1,1])
  }
})

patients <- do.call("rbind",patients)

mat <- ddply(patients,.(Patient),summarise,"Switches"=toString(GeneId))
mat$Switches <- gsub(", ","\t",mat$Switches)

write.table(mat,file=paste0(workingDir,"/switches/tables/",cancer,".patients.txt"),sep="\t", row.names=F, col.names=F, quote=F)

```